Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions

### 2023-04-05 StatPaper/APA/wileyNJD-APA.tex
updated text

\documentclass[APA,LATO1COL]{WileyNJD-v2}
\usepackage{lmodern}
\articletype{Article Type}%

\received{26 April 2016}
\revised{6 June 2016}
\accepted{6 June 2016}

@@

\presentaddress{This is sample for present address text this is sample for present address text}

\abstract[Summary]{Neural network-based survival methods can model data-driven
covariate interactions. While these methods have led to better
predictive performance than regression-based approaches, they cannot
model both time-varying interactions and complex baseline hazards. To
address this, we propose Case-Base Neural Networks (CBNN) as a new
approach that combines the case-base sampling framework with flexible
architectures. Our method naturally accounts for censoring and does not
require method specific hyperparameters. Using a novel sampling scheme
and data augmentation, we incorporate time directly into a feed-forward
neural network. CBNN predicts the probability of an event occurring at a
given moment and estimates the hazard function. We compare the
performance of CBNN to survival methods based on regression and neural
networks in a simulation and three real data applications. We report
two time-dependent metrics for each model. In the simulations and real
data applications, CBNN provides a more consistent predictive
performance across time and outperforms the competing neural network
approaches. For a complex simulation, which highlights the ability of
CBNN to model both a complex baseline hazard and time-varying
interactions, CBNN outperforms all competitors. The first two real data
application shows CBNN outperforming all neural network competitors,
while a second real data application shows competitive performance.
We highlight the benefit of combining case-base sampling with deep
learning to provide a simple and flexible modeling framework for
data-driven, time-varying interaction modeling of
survival outcomes. An R package is available at
\url{https://github.com/Jesse-Islam/cbnn}.}

@@

Smooth-in-time accelerated failure time (AFT) models can estimate
absolute risks by modeling the hazard directly through a user-specified
baseline hazard distribution \citep{kleinbaum2012survival}. Cox
proportional hazards models are used more often than AFT models, causing
analyses to be based on hazard ratios and relative risks rather than on
survival curves and absolute risks \citep{hanley2009}. The
identification of an appropriate distribution for the baseline hazard in
an AFT model may be difficult for common diseases that have many
interacting risk factors, or a Cox model where the disease pathogenesis
may change with age, making it difficult to maintain the proportional
hazards assumption. For example, previous studies of breast cancer
incidence have discovered time-varying interactions with covariates of
interest, such as tumor size \citep{coradini2000time}. One approach to
provide flexibility in the baseline hazard involves using the basis of
splines on time in our model \citep{royston2002flexible}. However,
regression-based models are limited in that they require prior knowledge
about potential time-varying interactions and their quantitative
effects.

Neural networks provide a data-driven approach to approximating
interaction terms. For example, DeepSurv is a neural network-based
proportional hazards model that implements the Cox partial
log-likelihood as a custom loss function \citep{katzman2018DeepSurv},
resulting in a stepwise absolute risk curve that cannot accommodate
time-varying interactions. Compared to Cox regression, DeepSurv shows
better performance on a real dataset \citep{katzman2018DeepSurv}. To handle
non-proportional hazards, a modification to the loss function was
proposed \citep{faraggi1995neural}.

As an alternative method that assumes a baseline hazard distribution,
DeepHit specifies each survival time of interest in the model and
directly estimates survival curves, rather than deriving a hazard
function . It assumes an inverse Gaussian
distribution as the baseline hazard and it outperformed DeepSurv \citep{lee2018DeepHit}.

Providing further flexibility, Deep Survival Machines (DSM) is a
parametric survival model using neural networks with a mixture of
distributions as the baseline hazard \citep{dsmPaper}. DSM outperformed DeepSurv and DeepHit
\citep{dsmPaper}. However, like DeepHit and DeepSurv, DSM cannot model
time-varying interactions.

We note that these alternative neural network approaches all require
custom loss functions \citep{katzman2018DeepSurv} \citep{lee2018DeepHit}
\citep{dsmPaper}. DeepHit introduces a hyperparameter weighing its two
loss functions (negative log-likelihood and ranking losses) while DSM
requires a two-phase learning process and user implementations for
distributions beyond Log-Normal or Weibull \citep{lee2018DeepHit}
\citep{dsmPaper}. Regression-based approaches require prior
specification of all interaction terms, which makes it challenging to
model covariate effects that change over time. The current neural
network models provide flexibility at the cost of opacity, while
regression models provide clarity at the cost of flexibility.

In this article, we propose Case-Base Neural Networks (CBNN) as a method
that models time-varying interactions and a flexible baseline hazard
using commonly available neural network components. Our approach to
modeling the full hazard uses case-base sampling \citep{hanley2009}.
This sampling technique allows probabilistic models to predict survival
outcomes. As part of the case-base framework, we use transformations of
time as a feature (covariate) to specify different baseline hazards. For
example, by including splines of time as covariates, we can approximate
the Royston-Parmar flexible baseline hazard model
\citep{royston2002flexible}, however, this still requires explicit use
of time-varying interactions. CBNN can model both without extra tuning
parameters.

In Section \ref{methods}, we describe how case-base sampling and neural
networks are combined both conceptually and algebraically, along with
our hyperparameter choices and software implementation. Section
\ref{sims}, describes our metrics and compares the performance of CBNN,
DeepSurv, DeepHit, DSM, Cox regression and case-base using logistic
regression (CBLR) on simulated data. Section \ref{casestudies} describes
the real-data analysis, while Section \ref{discussion} explores the
implications of our results and contextualizes them within neural
network survival analysis in a single event setting.

@@

\hypertarget{methods}{%
\section{Case-base neural network methodology, metrics and
software}\label{methods}}

In this section, we define case-base sampling, which converts the total
survival time into discrete person-specific moments (person-moments).
Then, we detail how neural networks can be used within this framework,
explicitly incorporating time as a feature while adjusting for the
sampling bias. Finally, we report on the software versions used. An R
package is available for use at
\url{https://github.com/Jesse-Islam/cbnn}. The entire code base to
reproduce the figures and empirical results in this paper is available
at \url{https://github.com/Jesse-Islam/cbnnManuscript}.

@@

Case-base sampling is an alternative framework for survival analysis
\citep{hanley2009}. In case-base sampling, we sample from the continuous
survival time of each person in our dataset to create a \emph{base
series} of \emph{person-moments}. This \emph{base series} complements
the \emph{case series}, which contains all person-moments at which the
event of interest occurs.

@@

Now, let \(b\) be the (user-defined) size of the \emph{base series} and
let \(B\) be the sum of all follow-up times for the individuals in the
study. If we sample the \emph{base series} uniformly across the study
base, then the hazard function of the sampling process is equal to
\(b/B\). Therefore, we have the following equality
\footnote{We are abusing notation here, conflating hazards with probabilities. For a rigorous treatment, see Saarela \& Hanley (2015) section 3 \cite{saarela2015} .}:
\begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align} The odds of a person-moment being a part of the \emph{case
series} is the ratio of the hazard \(h(T_i \mid X_i)\) and the uniform
rate \(b/B\). Using \eqref{eqn:main}, we can see how the log-hazard
function can be estimated from the log-odds arising from case-base
sampling: \begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

To estimate the correct hazard function, we adjusting for the bias
introduced when sampling a fraction of the study base \(B\) by adding
the offset term \(\log\left(\frac{b}{B} \right)\) as in
\eqref{eqn:offset}. Next, we propose using neural networks to model the
odds.

@@

After case-base sampling, we pass all features, including time, into any
user-defined feed-forward component, to which an offset term is added,
then passed through a sigmoid activation function (Figure
\ref{fig:NNarch}). We use the sigmoid function as its inverse is the
odds, which we can use to calculate the hazard. The general form for the
neural network using CBNN is:

\begin{align}\label{eqn:nnProb}
P\left(Y=1|X,T\right)&=\mathrm{sigmoid}\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) \right),
\end{align}

where \(T\) is a random variable representing the event time, \(X\) is
the random variable for a covariate profile, \(f_{\theta}(X, T)\)
represents any feed-forward neural network architecture,
\(\log\left(\frac{B}{b}\right)\) is bias term set by case-base sampling,
\(\theta\) is the set of parameters learned by the neural network and
\(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By approximating a
higher-order polynomial of time using a neural network, the baseline
hazard specification is now data-driven, where user-defined
hyperparameters such as regularization, number of layers and nodes
control the flexibility of the hazard function. We provide a detailed
description of the choices we made in the next sub-section.

@@

{\centering \includegraphics[width=1\linewidth]{../../../figures/nnarch2}}

@@

Backpropagation with an appropriate minimization algorithm (e.g.~Adam,
RMSPropagation, stochastic gradient descent) is used to optimize the
parameters in the model \citep{gulli2017}. For our analysis, we use Adam
as implemented in Keras \citep{gulli2017}. Note that the size of the
\emph{case series} is fixed as the number of events, but we can make the
\emph{base series} as large as we want. A ratio of 100:1 \emph{base
series} to \emph{case series} is sufficient \citep{hanley2009}. We pass
our feed-forward neural network through a sigmoid activation function
(Figure \ref{fig:NNarch}). Finally, we can convert this model output to
a hazard. When using our model for predictions, we manually set the
offset term to 0 in the new data, as we account for the bias during the
fitting process.

@@

@@

Neural networks are flexible when defining the architecture and
optimization parameters. These hyperparameter choices can affect the
estimated parameters and are not necessarily the same for each model.
First we fix a test set with 15\% of the data, then we use three-fold cross-validation on the remaining 
data (85\%:15\% training:validation) for a range of hyperparameters (TABLE REF). We track integrated brier 
score on the validation set for each hyperparameter combination, choosing 
the combination with the lowest score for each method. The chosen 
hyperparameters are in (TABLE REF).

@@

R \citep{Rsoft} and python \citep{py} are used to evaluate methods from
both languages. We fit the Cox model using the \textbf{survival} package
\citep{survpkg}, the CBLR model using the \textbf{casebase} package
\citep{cbpkg}, the DeepSurv model using the \textbf{survivalmodels}
package \citep{survmods}, the DeepHit model using \textbf{pycox}
\citep{lee2018DeepHit} and the DSM model using
\textbf{DeepSurvivalMachines} \citep{dsmPaper}. We made the components
of CBNN using the \textbf{casebase} package for the sampling step and
the \textbf{keras} \citep{keras} package for our neural network
architecture. The \textbf{simsurv} package \citep{simsurv} is used for
our simulation studies, while \textbf{flexsurv} \citep{flexsurv} is used
to fit a flexible baseline hazard using splines for our complex
simulation. We use the implementation of \(C_{IPCW}\) from the python
package \textbf{sksurv} \citep{sksurv}. The \textbf{riskRegression}
package \citep{riskRegression} is used to get the Index of Prediction
Accuracy (IPA metric). Both metrics are described in detail in the
following section. We modify the \textbf{riskRegression} package to be
used with any user supplied risk function \(F\). To ensure that both R
and Python-based models are running in unison on the same data through
our simulations and bootstrap, we use the \textbf{reticulate} package
\citep{reticulate}.

@@

\hypertarget{performance-metrics}{%
\section{Performance metrics}\label{performance-metrics}}

To choose our method-specific hyperparameters, we use the Integrated 
Brier Score (IBS) \citep{graf1999}. We use two metrics to assess the 
performance of the different methods of interest on testing data: 1) (IPA) 
\citep{kattan2018index} and 2) inverse probability
censoring weights-adjusted concordance index (\(C_{IPCW}\))
\citep{uno2011}, which we define below. IBS provides a summarized assessment
of performance for each model. (\(C_{IPCW}\)) provides information
about performance up to the survival time of interest. while IPA provide transparency
as to when in follow-up time each model may
perform better than the others.

The IPA is a function of the Brier score (\(BS(t)\)) \citep{graf1999},
which is defined as \begin{align}
BS(t)=\frac{1}{N}\sum^{N}_{i=1}\left(\frac{\left(1 - \widehat{F}(t \mid X_{i})\right)^{2}\cdot I(T_{i}\leq t,\delta_{i}=1)}{\widehat{G}(T_{i})} + \frac{\widehat{F}(t\mid X_{i})^{2}\cdot I(T_{i}>t)}{\widehat{G}(t)}\right),
\end{align} where \(\delta_{i}=1\) shows individuals who have
experienced the event, \(N\) represents the number of samples in our
dataset over which we calculate \(BS(t)\), \(\widehat{G}(t)=P[c>t]\) is
a non-parametric estimate of the censoring distribution, \(c\) is
censoring time and \(T_{i}\) is an individual's survival or censoring
time. The Brier score provides a score that accounts for the information
loss because of censoring. There are three categories of individuals
that may appear within the dataset once we fix our \(t\) of interest.
Individuals who experienced the event before \(t\) are present in the
first term of the equation. The second term of the equation includes
individuals who experience the event or are censored after \(t\). Those
censored before \(t\) are the third category of people. The inverse
probability censoring weights (IPCW) adjustment (\(G(\cdot)\)) is to
account for these category three individuals whose information is
missing. The IPA score as a function of time is given by \begin{align}
\textrm{IPA}(t) &= 1-\frac{BS_{model}(t)}{BS_{null}(t)}, \nonumber
\end{align} where \(BS_{model}(t)\) represents the Brier score over time
\(t\) for the model of interest and \(BS_{null}(t)\) represents the
Brier score if we use an unadjusted Kaplan-Meier (KM) curve as the
prediction for all observations \citep{kattan2018index}. Note that IPA
has an upper bound of one, where positive values show an increase in
performance over the null model and negative values show that the null
model performs better. These scores show how performance changes over
follow-up time.

A potential artifact of IPA is that the score is unstable at earlier and
later survival times. This is because of a near equivalent Brier score
among each model and the null model. At small values, a difference of
\(0.1\) creates a more significant fold change than at larger values. As
the Brier score is potentially small at the start and at the end of
their respective curves, The IPA score may be unstable at the same
locations.

The IBS is a function of the Brier score (\(BS(t)\)) \citep{graf1999}. as well,
which is defined as \begin{align}
IBS=\int_{t}^{t_{max}}BS(t)dw(t)
\end{align} where $w(t)=\frac{t}{t_{max}}$. As this is a score for a range of follow-up-time,
it is a useful metric to track during cross-validation. The performance at a specific time
is hidden, and is why we opt for the IPA in assessing the final models on a test set.

\hypertarget{inverse-probability-censoring-weights-adjusted-concordance-index}{%
\subsubsection{Inverse probability censoring weights-adjusted
concordance
index}\label{inverse-probability-censoring-weights-adjusted-concordance-index}}

The \(C_{IPCW}\) is a non-proper, rank-based metric that does not depend
on the censoring times in the test data \citep{uno2011}. The
\(C_{IPCW}\) is given by

\begin{align} \label{eq:cidx}
C_{IPCW}(t) &= \frac{\sum^{N}_{i=1}\sum^{N}_{j=1}\delta_{i}\left\{\widehat{G}(T_{i})\right\}^{-2} I(T_{i}<T_{j},T_{i}<t) I\left(\widehat{F}(t|X_{i})>\widehat{F}(t|X_{j})\right)}{\sum^{N}_{i=1}\sum^{N}_{j=1}\delta_{i}\left\{\widehat{G}(T_{i})\right\}^{-2} I(T_{i}<T_{j},T_{i}<t)}.
\end{align} where the follow-up period of interest is (0,\(t\)),
\(I(\cdot)\) is an indicator function, \(\widehat{F}(X_{i},t)\) is the
risk function estimated for everyone in the study at time \(t\) and
\(C_{IPCW}\) can compare the performance of different models, where a
higher score is better. Note that the \(C_{IPCW}\) may produce
misleading performance, as it ranks based on survival times, not event
status \citep{cindexfails2019}. This metric is considered an unbiased
population concordance measure because of the IPCW adjustment
\citep{uno2011}.

@@

{\centering \includegraphics[width=1\linewidth]{../../../analyses/figures/megaPlot}

@@

\caption{Summarizes the simple simulation (A, B), complex simulation (C, D), SUPPORT case study (E, F) and METABRIC case study (G, H) results. The first row shows the IPA score for each model in each study over follow-up time. Negative values mean our model performs worse than the null model and positive values mean the model performs better. The second row demonstrates the $C_{IPCW}$ score for each model in each study over follow-up time. A score of 1 is the maximum performance for either metric. Each model-specific metric in each study shows a 95-percent confidence interval over 100 iterations. The models of interest are case-base with logistic regression (CBLR), Case-Base Neural Networks (CBNN), Cox, DeepHit, DeepSurv, Deep Survival Machines (DSM), Optimal (a CBLR model with the exact interaction terms and baseline hazard specified) and Kaplan-Meier (to serve as a baseline, predicting the average for all individuals).}\label{fig:megaPlot}
\end{figure}

>>>>

\documentclass[APA,LATO1COL]{WileyNJD-v2}
\usepackage{lmodern}

\articletype{Article Type}%

\received{26 April 2023}
\revised{6 June 2023}
\accepted{6 June 2023}

@@

\presentaddress{McGill genome center: 740 Dr Penfield Ave, Montreal, Quebec H3A 0G1}

\abstract[Summary]{Neural network-based survival methods can model data-driven covariate interactions.
While these methods have led to better predictive performance than regression-based approaches,
they cannot model both time-varying interactions and complex baseline hazards. To address this, we
propose Case-Base Neural Networks (CBNN) as a new approach that combines the case-base sampling
framework with flexible architectures. Our method naturally accounts for censoring and does not require
method specific hyperparameters. Using a novel sampling scheme and data augmentation, we incorporate
time directly into a feed-forward neural network. CBNN predicts the probability of an event occurring at a
given moment to estimate the hazard function. We compare the performance of CBNN to survival methods
based on regression and neural networks in a simulation and three real data applications. We report two time-dependent
metrics for each model. For a complex simulation, which highlights the ability of CBNN to model both a complex
baseline hazard and time-varying interactions and two real data applications, CBNN outperforms all competitors
on a simulation with a complex baseline hazard and time-varying interaction and two case studies. A final case
study shows competitive performance. We highlight the benefit of combining case-base sampling with deep
learning to provide a simple and flexible modeling framework for data-driven, time-varying interaction modeling
of survival outcomes. An R package is available at https://github.com/Jesse-Islam/cbnn.}

@@

Smooth-in-time accelerated failure time (AFT) models can estimate absolute
risks by modeling the hazard directly through a user-specified baseline hazard
distribution \citep{kleinbaum2012survival}. Analysts use Cox proportional hazards
models more often than AFT models \citep{hanley2009}. This preference causes
analyses to be based on hazard ratios and relative risks rather than on survival
curves and absolute risks \citep{hanley2009}. With AFT models, it may be difficult
to identify an appropriate baseline hazard for common diseases with many
interacting risk factors \citep{royston2002flexible}. In studies where the disease
pathogenesis may change with age, we may see non-proportional hazards
\citep{coradini2000time}. For example, previous studies of breast cancer incidence
have discovered time-varying interactions with covariates of interest, such as tumor
size \citep{coradini2000time}. One approach to provide flexibility in the baseline hazard
involves using the basis of splines on time in the model \citep{royston2002flexible}.
However, regression-based models require prior knowledge of potential time-varying
interactions and their quantitative effects.

Neural networks provide a data-driven approach to approximating interaction terms.
For example, DeepSurv is a neural network-based proportional hazards model that
implements the Cox partial log-likelihood as a custom loss function \citep{katzman2018DeepSurv}.
This results in a stepwise absolute risk curve that cannot accommodate time-varying interactions.
Compared to Cox regression, DeepSurv shows better performance on a real dataset
\citep{katzman2018DeepSurv}. Previous work suggests we may change the loss function
to address non-proportional hazards \citep{faraggi1995neural}.

DeepHit assumes an inverse Gaussian distribution as the baseline hazard \citep{lee2018DeepHit}.
It specifies each follow-up time time of interest in the model and directly estimates survival curves, rather
than deriving a hazard function \citep{lee2018DeepHit}. DeepHit outperformed DeepSurv
\citep{lee2018DeepHit}.

We note that these alternative neural network approaches require custom loss functions
\citep{katzman2018DeepSurv} \citep{lee2018DeepHit}. DeepHit introduces a hyperparameter
weighing its two loss functions (negative log-likelihood and ranking losses) \citep{lee2018DeepHit}.
Regression-based approaches require prior specification of all interaction terms, which makes it
challenging to model covariate effects that change with time. The current neural network models
provide flexibility at the cost of clarity, while regression models provide clarity at the cost of flexibility.

In this article, we propose Case-Base Neural Networks (CBNN) as a method that models time-varying
interactions and a flexible baseline hazard. Our approach to modeling the full hazard uses case-base
sampling \citep{hanley2009}. This sampling technique allows probabilistic models to predict survival
outcomes. After case-base sampling, we can easily implement the model in any language using
common neural network components. As part of the case-base framework, we use transformations
of time as a feature (covariate) to specify different baseline hazards. For example, by including
splines of time as covariates, we can approximate the Royston-Parmar flexible baseline hazard model
\citep{royston2002flexible}\citep{hanley2009}. However, this still requires explicit use of time-varying
interactions. CBNN can model both without extra tuning parameters.

We describe how case-base sampling and neural networks combine
in Section \ref{methods}, along with our hyperparameter selection procedure, metrics and software implementation.
We explore the performance of CBNN, DeepSurv, DeepHit, Cox regression and case-base using logistic
regression (CBLR) on simulated data in Section \ref{sims}. Section \ref{casestudies} describes the
performance of the same methods in three case studies. Section \ref{discussion} explores the
implications of our results and contextualizes them within neural network survival analysis in a
single event setting.

@@

\hypertarget{methods}{%
\section{Case-base neural networks, metrics, hyperparameters and
software}\label{methods}}

In this section, we define case-base sampling, which converts the total
survival time into discrete person-specific moments (person-moments).
Next, we detail how neural networks can be used within this framework,
explicitly incorporating time as a feature while adjusting for the
sampling bias. Then, we describe the metrics and hyperparameter selection procedure.
Finally, we report on the software versions used. An R
package is available for use at
\url{https://github.com/Jesse-Islam/cbnn}. The entire code base to
reproduce the figures and empirical results in this paper is available
at \url{https://github.com/Jesse-Islam/cbnnManuscript}.

@@

Case-base sampling is an alternative framework for survival analysis
\citep{hanley2009}. Without this preliminary step, we cannot model survival
outcomes with probabilistic methods. In case-base sampling, we sample from the continuous
survival time of each person in our dataset to create a \emph{base
series} of \emph{person-moments}. This \emph{base series} complements
the \emph{case series}, which contains all person-moments at which the
event of interest occurs.

@@

Now, let \(b\) be the (user-defined) size of the \emph{base series} and
let \(B\) be the sum of all follow-up times for the individuals in the
study. let $c$ be the number of events in the \emph{case series}.
A reasonable concern is determining how large $b$ should be relative to $c$.
The size of \(b\) influences how much information is lost in the sampling process
\citep{hanley2009}. Using our notation, Hanley $\&$ Miettinen (2009)
adapt a quote from Mantel (1973): "By the reasoning that $\frac{cb}{c+b} [=(\frac{1}{c}+\frac{1}{b}]$
measures the relative information in a comparison of two averages based on
sample sizes of c and b respectively, we might expect by analogy, which would
of course not be exact in the present case, that this approach would result in
only a moderate loss of information. (The practicing statistician is generally
aware of this kind of thing. There is little to be gained by letting the size of
one series, b, become arbitrarily large if the size of the other series, c, must
remain fixed)" \citep{hanley2009} \citep{mantel1}. If $b=100c$, then we expect learned
weights to be at most one percent higher than if the entire study base $B$ was used,
as they are proportional to $\frac{1}{c}+\frac{1}{100c}$ rather than $\frac{1}{c}+\frac{1}{\infty c}$ \citep{hanley2009}.

If we sample the \emph{base series} uniformly across the study
base, then the hazard function of the sampling process is equal to
\(b/B\). Therefore, we have the following equality \cite{saarela2015}
\footnote{We are abusing notation here, conflating hazards with probabilities. For a rigorous treatment, see Saarela \& Hanley (2015) section 3 \cite{saarela2015} .}:
\begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align} The odds of a person-moment being a part of the \emph{case
series} is the ratio of the hazard \(h(T_i \mid X_i)\) and the uniform
rate \(b/B\). Using \eqref{eqn:main}, we can see how the log-hazard
function can be estimated from the log-odds arising from case-base
sampling: \begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

To estimate the correct hazard function, we adjust for the bias
introduced when sampling a fraction of the study base (\(\frac{b}{B}\)) by adding
the term \(\log\left(\frac{B}{b} \right)\) to offset \(\log\left(\frac{b}{B} \right)\)
during the fitting process. Next, we propose using neural networks to model the
odds.

@@

After case-base sampling, we pass all features, including time, into any
user-defined feed-forward component, to which an offset term is added,
then passed through a sigmoid activation function (Figure
\ref{fig:NNarch}). We are interested in predicting the odds. This makes the sigmoid
activation function ideal as it is the inverse of the odds. which we can use to calculate the hazard. The general form for the
neural network using CBNN is:

\begin{align*}
P\left(Y=1|X,T\right)&=\mathrm{sigmoid}\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) \right),
\end{align*}

where \(T\) is a random variable representing the event time, \(X\) is
the random variable for a covariate profile, \(f_{\theta}(X, T)\)
represents any feed-forward neural network architecture,
\(\log\left(\frac{B}{b}\right)\) is the offset term for adjust for the bias
(\(\log\left(\frac{b}{B}\right)\)) set by case-base sampling,
\(\theta\) is the set of parameters learned by the neural network and
\(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By approximating a
higher-order polynomial of time using a neural network, the baseline
hazard specification is now data-driven, where user-defined
hyperparameters such as regularization, number of layers and nodes
control the flexibility of the hazard function. We provide a detailed
description of the choices we made in the next sub-section.

@@

{\centering \includegraphics[width=1\linewidth]{../figures/nnarch2}}

@@

Backpropagation with an appropriate minimization algorithm (e.g.~Adam,
RMSPropagation, stochastic gradient descent) is used to optimize the
parameters in the model \citep{gulli2017}. For our analysis, we use Adam
as implemented in Keras \citep{gulli2017}. Note that the size of the
\emph{case series} is fixed as the number of events, but we can make the
\emph{base series} as large as we want. A ratio of 100:1 \emph{base
series} to \emph{case series} is sufficient \citep{hanley2009}. We pass
our feed-forward neural network through a sigmoid activation function
(Figure \ref{fig:NNarch}). Finally, we can convert this model output to
a hazard. To use CBNN for predictions, we manually set the
offset term \(\log\left(\frac{B}{b} \right)\) to 0 in the new data as we already account for the bias during the
fitting process.

@@

\hypertarget{performance-metrics}{%
\subsection{Performance metrics}\label{performance-metrics}}

To choose our method-specific hyperparameters, we use the Integrated
Brier Score (IBS) \citep{graf1999}. We use two metrics to assess the
performance of the different methods of interest on testing data: 1) Index of Prediction Accuracy (IPA)
\citep{kattan2018index} and 2) Inverse probability censoring weights-adjusted time dependent
area under the receiver operating characteristic curve ($AUC_{IPCW}$) \citep{auc}, which we 
define below. We achieve a summarized assessment of performance for each model with the
IBS. The IPA is a metric for both discrimination and calibration. The $AUC_{IPCW}$
provides a metric for discrimination only. Together, we can get a clearer picture about how
each model performs in each study and why.

The IPA is a function of the Brier score (\(BS(t)\)) \citep{graf1999},
which is defined as \begin{align}\label{eqn:bs}
BS(t)=\frac{1}{N}\sum^{N}_{i=1}\left(\frac{\left(1 - \widehat{F}(t \mid X_{i})\right)^{2}\cdot I(T_{i}\leq t,\delta_{i}=1)}{\widehat{G}(T_{i})} + \frac{\widehat{F}(t\mid X_{i})^{2}\cdot I(T_{i}>t)}{\widehat{G}(t)}\right),
\end{align} where \(\delta_{i}=1\) shows individuals who have
experienced the event, \(N\) represents the number of samples in our
dataset over which we calculate \(BS(t)\), \(\widehat{G}(t)=P[c>t]\) is
a non-parametric estimate of the censoring distribution, \(c\) is
censoring time and \(T_{i}\) is an individual's survival or censoring
time. The BS provides a score that accounts for the information
loss because of censoring. There are three categories of individuals
that may appear within the dataset once we fix our \(t\) of interest.
Individuals who experienced the event before \(t\) are present in the
first term of the equation. The second term of the equation includes
individuals who experience the event or are censored after \(t\). Those
censored before \(t\) are the third category of people. The inverse
probability censoring weights (IPCW) adjustment (\(G(\cdot)\)) is to
account for these category three individuals whose information is
missing.

The IPA as a function of time is given by \begin{align}
\textrm{IPA}(t) &= 1-\frac{BS_{model}(t)}{BS_{null}(t)}, \nonumber
\end{align} where \(BS_{model}(t)\) represents the BS over time
\(t\) for the model of interest and \(BS_{null}(t)\) represents the
BS if we use an unadjusted Kaplan-Meier (KM) curve as the
prediction for all observations \citep{kattan2018index}. Note that IPA
has an upper bound of one, where positive values show an increase in
performance over the null model and negative values show that the null
model performs better. These scores show how performance changes over
follow-up time. As an extension of BS, IPA models for both calibration
and discrimination\citep{graf1999} \citep{kattan2018index}.

The Integrated Brier Score (IBS) is a function of the BS \ref{eqn:bs} as well \citep{graf1999},
which is defined as \begin{align*}
IBS(t)=\int_{t}^{t_{max}}BS(t)dw(t),
\end{align*} where $w(t)=\frac{t}{t_{max}}$. As this is a score for a range of follow-up-times,
it is a useful metric to track during cross-validation. The performance at a specific time is lost,
and is why we opt for the IPA in assessing performance on a test set.

\hypertarget{auc}{%
\subsubsection{Inverse probability censoring weights-adjusted time dependent area under the receiver operating characteristic curve ($AUC_{IPCW}$) }\label{auc}}

The IPCW-adjusted AUC ($AUC_{IPCW}$) is a time-dependent metric that considers censoring \citep{auc}. 
For a given follow-up time of interest,
 \begin{align*}
AUC_{IPCW}(t)=\frac{\sum^{n}_{i=1}\sum^{n}_{j=1} I \left(\widehat{F}(t|X_{i})>\widehat{F}(t|X_{j})\right) \cdot I(T_{i}\leq t,\delta_{i}=1) \cdot \left( 1-I(T_{j}\leq t,\delta_{j}=1) \right) \cdot W_{i}(t) \cdot W_{j}(t)} { \sum^{n}_{i=1}\sum^{n}_{j=1} I(T_{i}\leq t,\delta_{i}=1) \cdot \left( 1-I(T_{j}\leq t,\delta_{j}=1) \right) \cdot W_{i}(t) \cdot W_{j}(t) },
\end{align*}where \begin{align*}
W_{i}(t)=\frac{I(T_{i}\leq t,\delta_{i}=1)}{\widehat{G}(T_{i})} +\frac{I(T_{i}>t)}{\widehat{G}(t)}.
\end{align*} The $AUC_{IPCW}$ measures discrimination \citep{auc}.
By taking both IPA and $AUC_{IPCW}$ together, we can better understand whether a model is better
in terms of calibration or discrimination.

@@

Neural networks are flexible when defining the architecture and optimization parameters.
These hyperparameter choices can affect the estimated parameters and are not necessarily
the same for each model. We apply the same procedure to each study in this paper. First,
we fix a test set with 15\% of the data. We keep aside this data during hyperparameter
selection and we fit a model for each method in each study. To determine the best
hyperparameters for each neural network method, we use three-fold cross-validated grid
search \citep{gulli2017} on the remaining data (85\% training, 15\% validation) for the
following range of hyperparameters:

\[
Grid\ search: \begin{cases}
Learning\ rate \sim \{0.001, 0.01\} \\ %technically a control and treatement group
Dropout \sim \{0.01,0.05, 0.1\} \\
First\ layer\ nodes \sim \{50, 75, 100\} \\
Second\ layer\ nodes \sim \{10,25,50\} \\
Number\ of\ batches \sim \{100, 500\} \\
Activation\ function \sim \{ReLU, Linear\}\\
\alpha \sim \{0, 0.5, 1\}.
\end{cases}
\]

\begin{table}
\caption{Four tables showing the chosen hyperparameters along with the IBS for each neural network model in the complex simulation (A), multiple myeloma (MM) case study (B), free light chain (FLC) case study (C) and prostate cancer (Prostate) case study (D).}
\label{tab:wins}
\begin{center}\includegraphics[width=1\linewidth]{../figures/winTable} \end{center}
\end{table}

DeepHit uses $\alpha$ as a specific hyperparameter, while DeepSurv and CBNN do not.
We track IBS on the validation set for each hyperparameter combination,
choosing the combination with the lowest score for each method.
We present the chosen hyperparameters in Table 1.

@@

R \citep{Rsoft} and python \citep{py} are used to evaluate methods from
both languages. We fit the Cox model using the \textbf{survival} package
\citep{survpkg}, the CBLR model using the \textbf{casebase} package
\citep{cbpkg}, DeepSurv and DeepHit using \textbf{pycox}
\citep{lee2018DeepHit}. We made the components
of CBNN using the \textbf{casebase} package for the sampling step and
the \textbf{keras} \citep{keras} package for our neural network
architecture. The \textbf{simsurv} package \citep{simsurv} is used for
our simulation studies, while \textbf{flexsurv} \citep{flexsurv} is used
to fit a flexible baseline hazard using splines for our complex
simulation. The \textbf{riskRegression}
package \citep{riskRegression} is used to get the Index of Prediction
Accuracy (IPA metric) and AUC. Both metrics are described in detail in the
following section. We modify the \textbf{riskRegression} package to be
used with any user supplied risk function \(F\). To ensure that both R
and Python-based models are running in unison on the same data through
our simulations and bootstrap, we use the \textbf{reticulate} package
\citep{reticulate}.

@@

@@

{\centering \includegraphics[width=1\linewidth]{../figures/megaPlot}

@@

\caption{Summarizes the complex simulation (A, E), multiple myeloma (MM) case study (B, F), free light chain (FLC) case study (C, G) and prostate cancer (Prostate) case study (D, H). The first row shows the IPA for each model in each study over follow-up time. Negative values mean our model performs worse than the null model and positive values mean the model performs better. The second row shows the $AUC_{IPCW}$ for each model in each study over follow-up time. Each model-specific metric in each study shows a 95\% confidence interval over 100 iterations. The models of interest are case-base with logistic regression (CBLR), Case-Base Neural Networks (CBNN), Cox, DeepHit and DeepSurv. Both CBLR and Cox have near identical performance, resulting in curves that are overlapped. The Kaplan-Meier (KM) model serves as a baseline, predicting the average curve for all individuals. The Optimal (a CBLR model with the exact interaction terms and baseline hazard specified) model shows the best performance we can expect on the simulated data. }\label{fig:megaPlot}
\end{figure}
